---
title: 'lonliness-AD' , following jupyter note MR_analysis.ipynb
author: 'Jennifer x.x.c'
date: '2026-08-10'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
#STEP2B-LD clumping
library(dplyr); library(remotes); library(ieugwasr);library(usethis)
#edit_r_environ()
data_path='/Users/apple/Documents/GitHub/personlig-datavitenskapsportef-lje/MR-Dementia/'
loneliness_gwas=read.csv(file.path(data_path, 'processed/loneliness_candidate_snps.csv'))
dim(loneliness_gwas)
head(loneliness_gwas)
#STEP2C-Instrument selection
clump_input=loneliness_gwas %>%
  transmute(
    rsid=SNP,
    pval=P_BOLT_LMM,
    id='loneliness')
head(clump_input)
clumped <- ld_clump(
  dat = clump_input,
  clump_kb = 10000,
  clump_r2 = 0.001,
  clump_p = 5e-8,
  pop = 'EUR') # 13 independent SNPs  left, 527 variants were removed because they were in LD with stronger variants or were absent from the EUR LD reference panel.
nrow(clumped)
clumped
clumped$rsid
setwd('/Users/apple/Documents/GitHub/personlig-datavitenskapsportef-lje/MR-Dementia/processed/')
loneliness_instruments=loneliness_gwas[loneliness_gwas$SNP %in% clumped$rsid,]
write.csv(loneliness_instruments, 'loneliness_instruments_13SNPs.csv', row.names=FALSE)
```
```{r}
#STEP3a: AD outcome GWAS
## so far, we have finished: G→Loneliness
## next, we need to finish: G→AD
## afterwardsm we can finish: Loneliness→AD
#STEP3b: Search candidate AD GWAS
library(TwoSampleMR)
ao=available_outcomes()
dim(ao)
ad_candidates <- ao[grepl('Alzheimer', ao$trait, ignore.case = TRUE),]
nrow(ad_candidates)
names(ad_candidates)
ad_table=ad_candidates[, c('id','trait','author','year','population','ncase','ncontrol','sample_size','study_design','consortium','pmid','note')]
ad_table=ad_table[order(ad_table$year, decreasing = TRUE),]
ad_display <- as.data.frame(ad_table[, c('id','trait','author','year','population','ncase','ncontrol','sample_size','study_design','consortium','pmid','note')])
View(ad_display)
#### phenoty: Alzheimer's disease/late-onset Alzheimer's disease, exclude: parental history、maternal、paternal、family history、proxy/broad dementia
#### population: european 
#### case definition: clinically diagnosed/autopsy-confirmed AD
#### sample size/cases: as more as better, as the above tow criteria satisfy
#### data independency 
#### data quality and usability
ad_auto <- ad_table[ad_table$population == 'European' & !grepl( 'parent|maternal|paternal|family|proxy|history|dementia', ad_table$trait, ignore.case = TRUE),]
ad_auto <- ad_auto[order(ad_auto$ncase, decreasing = TRUE, na.last = TRUE),]
ad_auto[, c('id', 'trait', 'author', 'year','population', 'ncase', 'ncontrol', 'sample_size', 'pmid')]
View(ad_auto)
ad_auto2 <- ad_auto[grepl('Alzheimer|late onset',ad_auto$trait,ignore.case = TRUE),]
View(ad_auto2)
#STEP3c:  Verify phenotype / ancestry / sample
#STEP3b: Choose primary outcome
#STEP3b: Extract the 13 SNPs
ad_outcome=extract_outcome_data(snps = clumped$rsid, outcomes='ieu-b-2', proxies = FALSE) #Kunkle 2019
nrow(ad_outcome)
ad_outcome[, c('SNP', 'beta.outcome', 'se.outcome', 'effect_allele.outcome', 'other_allele.outcome', 'pval.outcome')]
table(ad_outcome$proxy.outcome)
View(ad_outcome)
setdiff(ad_outcome$SNP,clumped$rsid) # 5 SNPs in ad_outcome_direct are not within clumped, check the problem: rsID changed, but chromosome + position same, have to make them matched
ad_outcome$SNP_original <- ad_outcome$SNP
loneliness_instruments$key <- paste(loneliness_instruments$CHR, loneliness_instruments$BP,sep = ':')
ad_outcome$key <- paste(ad_outcome$chr,ad_outcome$pos,sep = ':')
snp_map <- setNames(loneliness_instruments$SNP,loneliness_instruments$key)
ad_outcome$SNP <- snp_map[ad_outcome$key]
setdiff(ad_outcome$SNP,loneliness_instruments$SNP) 
```
```{r}
#STEP4: Harmonisation: to make effect of exposure and outcome towards same allele
exposure_data <- data.frame(
  SNP = loneliness_instruments$SNP,
  beta.exposure = loneliness_instruments$BETA,
  se.exposure = loneliness_instruments$SE,
  effect_allele.exposure = loneliness_instruments$ALLELE1,
  other_allele.exposure = loneliness_instruments$ALLELE0,
  eaf.exposure = loneliness_instruments$A1FREQ,
  pval.exposure = loneliness_instruments$P_BOLT_LMM,
  exposure = 'Loneliness',
  id.exposure = 'Day2018')
harmonised <- harmonise_data(exposure_dat = exposure_data, outcome_dat = ad_outcome, action = 2)
```
```{r}
#STEP5: MR
mr_data <- harmonised[harmonised$mr_keep == TRUE,]
mr_results <- mr(mr_data)
mr_results
##IVW: Inverse variance weighted (IVW), (Beta_SNP->AD)/(Beta_SNP->Loneliness), effect estimate of Loneliness -> AD
##Weighted median: sensitivity method
##MR-Egger
```
```{r}
#STEP6: Sensitivity analyses
mr_or <- generate_odds_ratios(mr_results)
heterogeneity <- mr_heterogeneity(mr_data)
pleiotropy <- mr_pleiotropy_test(mr_data)
loo <- mr_leaveoneout(mr_data) #删掉某一个 SNP 后，总体IVW 会不会大变？
single <- mr_singlesnp(mr_data) #每一个 SNP 单独给出的 causal estimate 是多少？
p_scatter <- mr_scatter_plot(mr_results, mr_data)
```
#STEP7: Interpretation and reporting
Primary IVW: 
OR=0.64, 95%CI=0.1-4.01, P=0.631
There was no evidence of a causal association between genetically predicted liability to loneliness and clinically diagnosed Alzheimer's disease. However, the estimate was imprecise, with a wide confidence interval.
Sensitivity results：
Weighted median：OR=0.52,P=0.593
Weighted mode：OR=0.27,P=0.470
Simple mode：OR=0.19,P=0.442
MR-Egger：OR=1466,P=0.133
Heterogeneity：Q_IVW=10.60,P=0.389， no evidence of significant heterogeneity.
MR-Egger intercept：−0.043,P=0.107，no evidence of directional horizontal pleiotropy.

Leave-one-out plot: no single SNP appeared to drive the IVW estimate.

This two-sample MR analysis did not provide evidence that genetically predicted liability to self-reported loneliness causally influences the risk of clinically diagnosed Alzheimer's disease. The primary IVW estimate was imprecise, and sensitivity analyses did not indicate substantial heterogeneity, directional pleiotropy, or dominance by a single genetic instrument.
The wide confidence intervals indicate limited statistical precision, so the findings should be interpreted as inconclusive rather than evidence for the absence of a causal effect.


